Unifying intraspecific and interspecific variation in tropical tree mortality

### 2016-06-14 ms/ms.Rnw
Standardise census indices throughout
- opting for t_1, t_2, t_3

Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions. In our dataset, \Sexpr{proportion_negative}\% of observed growth rates were negative. While there are physiological processes (e.g. water depletion and bark drop) that can lead to trunk shrinkages overtime \cite{Pastur:2007kl}, it is biologically impossible for growth (i.e. the creation of new biomass) to be negative \cite{Clark:2007ve}. To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e. observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors \cite{Condit:2012nz}. This dataset consisted of \Sexpr{n_obs_errors} randomly selected trees that had their diameter measured twice within a 30 day period \cite{Condit:2012nz}. Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error. The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter. Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual. We did this by first modelling the observed initial ($D_{i,0}$) and final diameter ($D_{i,1}$) of each individual, $i$, as random realisations from a normal distribution centerd on their respective true estimates (i.e. $\widehat{D_{i,0}}$ and $\widehat{D_{i,1}}$), with the standard deviation fixed at \Sexpr{obs_error_sigma} (i.e. the observed measurement error):

\begin{equation}\label{eq:true_growth_obs} 
\begin{aligned}
D_{i,0} &\sim \mathcal{N}(\widehat{D}_{i,0}, 0.75) \\
D_{i,1} &\sim \mathcal{N}(\widehat{D}_{i,1}, 0.75)
\end{aligned} 
\end{equation}

The initial diameter for each individual $i$, $\widehat{D}_{i,0}$, was sampled as a random realisation from a log-normal distribution:
\begin{equation} \label{eq:true_dbh1}
\widehat{D}_{i,0} \sim \log \mathcal{N}(\mu_{D_0}, \sigma_{D_0}),
\end{equation}
where $\mu_{D_0}$ is the expected mean of log initial diameter and $\sigma_{D_0}$ defines that variation around that mean. $\mu_{D_0}$ was estimated using a normal prior $\mathcal{N}(0, 2.5)$, which represents our prior expectation that the mean initial diameter would be greater than zero but less than 150 cm. $\sigma_{D_0}$ was estimated using a weakly informative half Cauchy prior centerd on zero and a scale parameter of 2.5.

The final diameter measurement for each individual $i$, $D_{i,1}$ was estimated as:
\begin{equation}\label{eq:true_dbh2}
\widehat{D}_{i,1} = \widehat{D}_{i,0} + \Delta D_{i,1} \, (t_1-t_0),
\end{equation}
where $\Delta D_{i,1}$ represents the annual true increase in diameter (cm) and $(t_1-t_0)$ represents the number of days between initial and final measurements divided by 365.25. We modelled $\Delta D_{i,1}$ as a random realisation from a log-normal distribution centerd on zero and a standard deviation of 1. This prior conveyed our expectation that annual diameter growth could not be negative and is unlikely to be greater than 7 cm/yr.

Using this model we then used \texttt{rstan}'s \textit{optimizing} function to maximize the joint posterior and extract point estimates of true initial diameter, $\widehat{D}_{i,0}$, true final diameter, $\widehat{D}_{i,1}$, and true annual diameter growth $\Delta D_{i,1}$. We then used annual diameter growth, $\Delta D_{i,1}$, and annual basal area growth: $\frac{(\pi \, 0.25 \,\widehat{D}_{i,1}) - (\pi \, 0.25 \, \widehat{D}_{i,0})}{(t_1-t_0)}$ in the subsequent mortality analysis.

@@

In survival analysis, the probability that an individual dies between $t_1$ and $t_2$, $p_{t_1\rightarrow t_2}$ is defined by:
\begin{equation} \label{eq:mortality_model}
p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
\end{equation}
where, $\lambda$ is the instantaneous hazard rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_1}^{t_2} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_1$ and $t_2$.

@@

Our observations are for survival outcomes $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_1$ and $t_2$. We model these observations as random realizations from a Bernoulli distribution
\begin{equation}
S_{i,t_1\rightarrow t_2} \sim \textrm{Bernoulli}(p_{i,t_1\rightarrow t_2}),
\end{equation}
where $p_{i,t_1\rightarrow t_2}$ is the probability individual $i$ dies in the interval $t_1$--$t_2$. Using standard survival analysis, this probability is given by
\begin{equation}
p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
\end{equation}
As we only have a point estimate of $\lambda_i(t)$ at $t=t_1$, the cumulative hazard from  $t_1$--$t_2$ is approximated as
\begin{equation}
  \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t \approx \lambda_i(t_1) (t_2-t_1).
\end{equation}

@@

Predictive accuracy was quantified by estimating the average log loss across K-folds for held-out data, $\bar{\mathcal{L}}$ (Fig. \ref{fig1}). Logarithmic loss, also known as log loss and indicated by $\mathcal{L}$, measures the accuracy of a model by incorporating probabilistic confidence. It does this by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome (Fig ??). By minimizing the $\bar{\mathcal{L}}$, we maximize predictive skill. The average log loss across all individuals for the $k$th fold of held-out data, $\mathcal{L}_k$, is given by
\begin{equation} \label{eq:logloss}
\mathcal{L}_k = - \frac1{N_k}\,\sum_{i=1}^{N_k}\big(S_i\,\log(p_{i,t_1\rightarrow t_2}) + ( 1 - S_i)\,\log(1 - p_{i,t_1\rightarrow t_2})\big),
\end{equation}
where $N_k$ refers to the number of observations in fold.

>>>>

Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions. In our dataset, \Sexpr{proportion_negative}\% of observed growth rates were negative. While there are physiological processes (e.g. water depletion and bark drop) that can lead to trunk shrinkages overtime \cite{Pastur:2007kl}, it is biologically impossible for growth (i.e. the creation of new biomass) to be negative \cite{Clark:2007ve}. To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e. observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors \cite{Condit:2012nz}. This dataset consisted of \Sexpr{n_obs_errors} randomly selected trees that had their diameter measured twice within a 30 day period \cite{Condit:2012nz}. Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error. The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter. Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual. We did this by first modelling the observed initial ($D_{i,1}$) and final diameter ($D_{i,2}$) of each individual, $i$, as random realisations from a normal distribution centerd on their respective true estimates (i.e. $\widehat{D_{i,1}}$ and $\widehat{D_{i,2}}$), with the standard deviation fixed at \Sexpr{obs_error_sigma} (i.e. the observed measurement error):

\begin{equation}\label{eq:true_growth_obs} 
\begin{aligned}
D_{i,1} &\sim \mathcal{N}(\widehat{D}_{i,1}, 0.75) \\
D_{i,2} &\sim \mathcal{N}(\widehat{D}_{i,2}, 0.75)
\end{aligned} 
\end{equation}

The initial diameter for each individual $i$, $\widehat{D}_{i,1}$, was sampled as a random realisation from a log-normal distribution:
\begin{equation} \label{eq:true_dbh1}
\widehat{D}_{i,1} \sim \log \mathcal{N}(\mu_{D_1}, \sigma_{D_1}),
\end{equation}
where $\mu_{D_1}$ is the expected mean of log initial diameter and $\sigma_{D_1}$ defines that variation around that mean. $\mu_{D_1}$ was estimated using a normal prior $\mathcal{N}(0, 2.5)$, which represents our prior expectation that the mean initial diameter would be greater than zero but less than 150 cm. $\sigma_{D_1}$ was estimated using a weakly informative half Cauchy prior centerd on zero and a scale parameter of 2.5.

The final diameter measurement for each individual $i$, $D_{i,2}$ was estimated as:
\begin{equation}\label{eq:true_dbh2}
\widehat{D}_{i,2} = \widehat{D}_{i,1} + \Delta D_{i,2} \, (t_2-t_1),
\end{equation}
where $\Delta D_{i,2}$ represents the annual true increase in diameter (cm) and $(t_2-t_1)$ represents the number of days between initial and final measurements divided by 365.25. We modelled $\Delta D_{i,2}$ as a random realisation from a log-normal distribution centerd on zero and a standard deviation of 1. This prior conveyed our expectation that annual diameter growth could not be negative and is unlikely to be greater than 7 cm/yr.

Using this model we then used \texttt{rstan}'s \textit{optimizing} function to maximize the joint posterior and extract point estimates of true initial diameter, $\widehat{D}_{i,1}$, true final diameter, $\widehat{D}_{i,2}$, and true annual diameter growth $\Delta D_{i,2}$. We then used annual diameter growth, $\Delta D_{i,2}$, and annual basal area growth: $\frac{(\pi \, 0.25 \,\widehat{D}_{i,2}) - (\pi \, 0.25 \, \widehat{D}_{i,1})}{(t_2-t_1)}$ in the subsequent mortality analysis.

@@

In survival analysis, the probability that an individual dies between $t_2$ and $t_3$, $p_{t_2\rightarrow t_3}$ is defined by:
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
where, $\lambda$ is the instantaneous hazard rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_2$ and $t_3$.

@@

Our observations are for survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$. We model these observations as random realizations from a Bernoulli distribution
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}),
\end{equation}
where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$. Using standard survival analysis, this probability is given by
\begin{equation}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
As we only have a point estimate of $\lambda_i(t)$ at $t=t_2$, the cumulative hazard from  $t_2$--$t_3$ is approximated as
\begin{equation}
  \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t \approx \lambda_i(t_2) (t_3-t_2).
\end{equation}

@@

Predictive accuracy was quantified by estimating the average log loss across K-folds for held-out data, $\bar{\mathcal{L}}$ (Fig. \ref{fig1}). Logarithmic loss -- commonly known as log loss, and indicated by $\mathcal{L}$-- measures the accuracy of a model by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome (Fig ??). By minimizing the $\bar{\mathcal{L}}$, we maximize predictive skill. The average log loss across all individuals for the $k$th fold of held-out data, $\mathcal{L}_k$, is given by
\begin{equation} \label{eq:logloss}
\mathcal{L}_k = - \frac1{N_k}\,\sum_{i=1}^{N_k}\big(S_i\,\log(p_{i,t_2\rightarrow t_3}) + ( 1 - S_i)\,\log(1 - p_{i,t_2\rightarrow t_3})\big),
\end{equation}
where $N_k$ refers to the number of observations in fold.
